---
title: "PA2_template"
author: "carlos rios"
date: "Wednesday, August 20, 2014"
output: html_document
---
## Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.
In this work we aim to show  which types of events are most harmful with respect to population health and which types of events have the greatest economic consequences in United States. To make this work we obteined data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.We specifically work with data for the years from 1950 to 2011.

## Data Processing
blabla

### load libraries
```{r warning=FALSE,message=FALSE}
library("knitr")
library("plyr")
library("ggplot2")
library("lubridate")
library("stringr")
source("R code/dataProcessing.R")
```
### load data set

```{r cache=TRUE}
stormData<-read.csv("Data/repdata-data-StormData.csv")

```
### Processing data set
```{r}
stormData<-dataProcessing(stormData)
```

##Results 
blabla

```{r}
totalFatalities<-ddply(stormData,.(EVTYPE),summarise,totalFatalities=sum(FATALITIES))
totalFatalities<-totalFatalities[!(totalFatalities$totalFatalities==0),]
totalFatalities$EVTYPE<-as.factor(totalFatalities$EVTYPE)
i<-order(totalFatalities$totalFatalities,decreasing = T )
totalFatalities<-totalFatalities[i,]
```

```{r}
top20<-totalFatalities[1:20,]
top20$EVTYPE<- reorder(top20$EVTYPE, top20$totalFatalities,desc)
bar <- ggplot(top20, aes(y=totalFatalities))
bar + geom_bar(aes(x=EVTYPE),stat ="identity",binwidth=1 )+labs(title="20 first Event type and injuries",y="total injuries",x="")+theme(axis.text.x = element_text(hjust=1,angle = 45))

```

